---
title: '8\. Worksheet: Among Site (Beta) Diversity -- Part 1'

author: 'Eli Graber; Z620: Quantitative Biodiversity, Indiana University'
output: pdf_document
geometry: margin=2.54cm 

---

## OVERVIEW

In this worksheet, we move beyond the investigation of within-site $\alpha$-diversity.
We will explore $\beta$-diversity, which is defined as the diversity that occurs among sites.
This requires that we examine the compositional similarity of assemblages that vary in space or time.

After completing this exercise you will know how to:

1. formally quantify $\beta$-diversity
2. visualize $\beta$-diversity with heatmaps, cluster analysis, and ordination
3. test hypotheses about $\beta$-diversity using multivariate statistics

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the  worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, it is *imperative* that you **push** this file to your GitHub repo, at whatever stage you are. Ths will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your '8.BetaDiversity' folder.
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file 
(**8.BetaDiversity_1_Worksheet.Rmd**) with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**8.BetaDiversity_1_Worksheet.pdf**).

The completed exercise is due on **Friday, April 16^th^, 2021 before 09:00 AM**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/8.BetaDiversity*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}


rm=(list=ls())
getwd()
setwd("~/GitHub/QB2021_Graber/2.Worksheets/8.BetaDiversity")


```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `doubs` dataset from the `ade4` package, and
2. explore the structure of the dataset.

```{r}
# note, pleae do not print the dataset when submitting

package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies')
  for (package in package.list) {
    if (!require(package, character.only = TRUE, quietly = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
data(doubs)
str(doubs, max.level=1)
head(doubs$env)
package.list <- c('vegan', 'ade4','viridis', 'gplots', 'BiodiversityR', 'indicspecies')
for(package in package.list){
  if(!require(package, character.only = T, quietly = T)) {
    install.packages(package)
    library(package, character.only = T)
  }
}
data(doubs)

str(doubs, max.level = 1)
head(doubs$env)

```

***Question 1***:  Describe some of the attributes of the `doubs` dataset.  

a.  How many objects are in `doubs`?
b.  How many fish species are there in the `doubs` dataset?
c.  How many sites are in the `doubs` dataset?


> ***Answer 1a***:  4 lists
> ***Answer 1b***:  27 fish species
> ***Answer 1c***:  30 sites


### Visualizing the Doubs River Dataset

***Question 2***: Answer the following questions based on the spatial patterns of richness (i.e., $\alpha$-diversity) and Brown Trout (*Salmo trutta*) abundance in the Doubs River.  

a.  How does fish richness vary along the sampled reach of the Doubs River?
b.  How does Brown Trout (*Salmo trutta*) abundance vary along the sampled reach of the Doubs River?
c.  What do these patterns say about the limitations of using richness when examining patterns of biodiversity?


> ***Answer 2a***:  There is a high density of fish upstream and around the bend.
> ***Answer 2b***:  Brown Trout are particularly dense upstream and at the start of the river bend. 
> ***Answer 2c***: Considering richness of the overall area can take away from understanding the distribution of specific species.  
=======
> ***Answer 2a***:  
> ***Answer 2b***:  
> ***Answer 2c***:  


## 3) QUANTIFYING BETA-DIVERSITY

In the R code chunk below, do the following:  

1. write a function (`beta.w()`) to calculate Whittaker's $\beta$-diversity (i.e., $\beta_{w}$) that accepts a site-by-species matrix with optional arguments to specify pairwise turnover between two sites, and
2. use this function to analyze various aspects of $\beta$-diversity in the Doubs River.  

```{r}

beta.w <- function (site.by.species = "") {
  SbyS.pa <- decostand(site.by.species, method = "pa")
  S <- ncol(SbyS.pa[,which(colSums(SbyS.pa)>0)])
  a.bar <- mean(specnumber(SbyS.pa))
  b.w <- round(S/a.bar, 3)
  return(b.w)
}
beta.w <-function(site.by.species = "", sitenum1= "", sitenum2 ="", pairwise = FALSE) {
  if (pairwise == TRUE){
    if (sitenum1 == "" | sitenum2 == "") {
      print("Error: please specify sites to compare")
      return(NA) }
    site1 = site.by.species[sitenum1,]
    site2 = site.by.species[sitenum2,]
    site1 = subset(site1, select = site1 > 0)
    site2 = subset(site2, select = site2 > 0)
    gamma = union(colnames(site1), colnames(site2))
    s = length(gamma)
    a.bar = mean(c(specnumber(site1), specnumber(site2)))
    b.w =round(s/a.bar - 1, 3)
    return(b.w)
  }
else{
  SbyS.pa <- decostand(site.by.species, method = "pa")
  S <- ncol(SbyS.pa[,which(colSums(SbyS.pa) > 0)])
  a.bar <- mean(specnumber(SbyS.pa))
  b.w <- round(S/a.bar, 3)
  return(b.w)
  }
}
beta.w(doubs$fish)
beta.w(doubs$fish, 1, 2, pairwise = TRUE)
beta.w(doubs$fish, 1, 10, pairwise = TRUE)


```

***Question 3***: Using your `beta.w()` function above, answer the following questions:

a.  Describe how local richness ($\alpha$) and turnover ($\beta$) contribute to regional ($\gamma$) fish diversity in the Doubs.
b.  Is the fish assemblage at site 1 more similar to the one at site 2 or site 10?
c.  Using your understanding of the equation $\beta_{w} = \gamma/\alpha$, how would your interpretation of $\beta$ change if we instead defined beta additively (i.e., $\beta = \gamma - \alpha$)?


> ***Answer 3a***: Beta diversity output is 2.16 meaning that the local richness is actively (2.16 times) lower than regional thus there is relatively high turnover. 
> ***Answer 3b***:  The higher the output the higher the beta diversity meaning the more similar they are. Sites 1 and 10 have a higher out put (0.714) to site 1 to 2 (0.5) meaning taht sites 1 and 10 are more similar. 
> ***Answer 3c***:  We would get a value that is not comprable to other sites. The way beta diveristy is calculed allows it to be a ratio and thus comprable while subtraction would not allow that. 


### The Resemblance Matrix
In order to quantify $\beta$-diversity for more than two samples, we need to introduce a new primary ecological data structure: the **Resemblance Matrix**. 

***Question 4***: How do incidence- and abundance-based metrics differ in their treatment of rare species?


> ***Answer 4***: Incidence would over estimate the presence of rare species as it would just acknowledge if they do or do not occur in a certain area while abundance based would  acknowledge the frequency of their presence. 
=======
> ***Answer 4***:  


In the R code chunk below, do the following:  

1. make a new object, `fish`, containing the fish abundance data for the Doubs River, 
2. remove any sites where no fish were observed (i.e., rows with sum of zero),
3. construct a resemblance matrix based on Sørensen's Similarity ("fish.ds"), and
4. construct a resemblance matrix based on Bray-Curtis Distance ("fish.db").  

```{r}



fish <- doubs$fish
fish <- fish[-8,]
fish.dj <- vegdist(fish, method= "jaccard", binary = TRUE)
fish.db <- vegdist(fish, method = "bray")
fish.ds <- vegdist(fish, method = "bray", binary = TRUE, diag=TRUE)
fish.db <- vegdist(fish, method = "bray", upper = TRUE, diag = TRUE)
print(fish.db)
print(fish.ds)


```

***Question 5***: Using the distance matrices from above, answer the following questions:

a. Does the resemblance matrix (`fish.db`) represent similarity or dissimilarity? What information in the resemblance matrix led you to arrive at your answer? 
b. Compare the resemblance matrices (`fish.db` or `fish.ds`) you just created. How does the choice of the Sørensen or Bray-Curtis distance influence your interpretation of site (dis)similarity?


> ***Answer 5a***:  It represents dissimilarity because if we look at the the 1 compared to 1 output is zero (same for 2 to 2 etc.) so any number above 0 represents dissimilarities between those sites. 
> ***Answer 5b***: Sorensen has has a slightly lower output compared ot Bray-Curtis so interpretation of Bray-Curtis would lead to one thinking dissimilarity is higher than if you used Sorensen. 
=======
> ***Answer 5a***:  
> ***Answer 5b***:


## 4) VISUALIZING BETA-DIVERSITY

### A. Heatmaps

In the R code chunk below, do the following:  

1. define a color palette,
2. define the order of sites in the Doubs River, and
3. use the `levelplot()` function to create a heatmap of fish abundances in the Doubs River.

```{r}


order <- rev(attr(fish.db, "Labels"))
levelplot(as.matrix(fish.db)[, order], aspect = "iso", col.regions = inferno,
          xlab= "Doubs Site", ylab= "Doubs Site", scales = list(cex=0.5),
          main = "Bray-Curtis Distance")





```

### B.  Cluster Analysis

In the R code chunk below, do the following:  

1. perform a cluster analysis using Ward's Clustering, and
2. plot your cluster analysis (use either `hclust` or `heatmap.2`). 

```{r} 






fish.ward <- hclust(fish.db, method = "ward.D2")
par(mar = c(1,5,2,2) + 0.1)
plot(fish.ward, main= "Doubs River Fish: Ward's Clustering", 
     ylab= "Squared Bray-Curtis Distance")



```

***Question 6***: Based on cluster analyses and the introductory plots that we generated after loading the data, develop an ecological hypothesis for fish diversity the `doubs` data set? 


> ***Answer 6***: Generally speaking the closer the sites are together the more similar (the closeness of site number represents their closens geography) they will be with some exceptions to that rule such as 1, 5, 9, and 23-25. 


### C.  Ordination

### Principal Coordinates Analysis (PCoA)

In the R code chunk below, do the following:  

1. perform a Principal Coordinates Analysis to visualize beta-diversity
2. calculate the variation explained by the first three axes in your ordination
3. plot the PCoA ordination,
4. label the sites as points using the Doubs River site number, and
5. identify influential species and add species coordinates to PCoA plot.

```{r}


fish.pcoa <- cmdscale(fish.db, eig = TRUE, k = 3)
explainvar1 <-round(fish.pcoa$eig[1] / sum(fish.pcoa$eig), 3) *100
explainvar2 <-round(fish.pcoa$eig[2] / sum(fish.pcoa$eig), 3) *100
explainvar3 <-round(fish.pcoa$eig[3] / sum(fish.pcoa$eig), 3) *100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

par(mar= c(5,5,1,2)+0.1)
plot(fish.pcoa$points[,1], fish.pcoa$points[,2], ylim = c(-0.2,0.7),
     xlab= paste("PCoA 1(", explainvar1, "%)", sep = ""),
     ylab= paste("PCoA 2(", explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis =1.2, axes=FALSE)
axis(side = 1, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
axis(side = 2, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
abline(h = 0, v = 0, lty =3)
box(lwd = 2)

points(fish.pcoa$points[,1], fish.pcoa$points[,2],
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[,1], fish.pcoa$points[,2],
     labels = row.names(fish.pcoa$points))
fishREL <- fish
  for(i in 1:nrow(fish)){
    fishREL[i,] = fish[i,]/ sum(fish[i,])
  }
fish.pcoa <- add.spec.scores(fish.pcoa,fishREL, method = "pcoa.scores")
text(fish.pcoa$cproj[,1], fish.pcoa$cproj[,2],
     labels = row.names(fish.pcoa$cproj), col = "black")
=======

```

In the R code chunk below, do the following:  

1. identify influential species based on correlations along each PCoA axis (use a cutoff of 0.70), and
2. use a permutation test (999 permutations) to test the correlations of each species along each axis.

```{r}


fish.pcoa <- cmdscale(fish.db, eig = TRUE, k = 3)
explainvar1 <-round(fish.pcoa$eig[1] / sum(fish.pcoa$eig), 3) *100
explainvar2 <-round(fish.pcoa$eig[2] / sum(fish.pcoa$eig), 3) *100
explainvar3 <-round(fish.pcoa$eig[3] / sum(fish.pcoa$eig), 3) *100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

par(mar= c(5,5,1,2)+0.1)
plot(fish.pcoa$points[,1], fish.pcoa$points[,2], ylim = c(-0.2,0.7),
     xlab= paste("PCoA 1(", explainvar1, "%)", sep = ""),
     ylab= paste("PCoA 2(", explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis =1.2, axes=FALSE)
axis(side = 1, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
axis(side = 2, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
abline(h = 0, v = 0, lty =3)
box(lwd = 2)

points(fish.pcoa$points[,1], fish.pcoa$points[,2],
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[,1], fish.pcoa$points[,2],
     labels = row.names(fish.pcoa$points))
fishREL <- fish
  for(i in 1:nrow(fish)){
    fishREL[i,] = fish[i,]/ sum(fish[i,])
  }
fish.pcoa <- add.spec.scores(fish.pcoa,fishREL, method = "pcoa.scores")
text(fish.pcoa$cproj[,1], fish.pcoa$cproj[,2],
     labels = row.names(fish.pcoa$cproj), col = "black")
spe.corr <- add.spec.scores(fish.pcoa, fishREL, method = "cor.scores")$cproj
corrcut <- 0.7
imp.spp <- spe.corr [abs(spe.corr[,1]) >= corrcut | abs(spe.corr[,2]) >= corrcut, ]
fit <- envfit(fish.pcoa, fishREL, perm = 999)
=======






```

***Question 7***: Address the following questions about the ordination results of the `doubs` data set:

a.  Describe the grouping of sites in the Doubs River based on fish community composition.
b.  Generate a hypothesis about which fish species are potential indicators of river quality.


> ***Answer 7a***:  Sites 16-19 have a large variety of fish speices, there aren't fish that are particularly common in 20:22,26:30, while 1:15 (except 9) are relatively similar with few common fish (Satr, Neba, Phph). 
> ***Answer 7b***:   The high density of such a variety of fish in the sites 16-19 is representative of a high quality enviornment becuase such a diversity of fish being maintianed while the presence of Satr in the 1:15 area shows it being lower quality which may also be true of 23:25 having Alal. 



## SYNTHESIS

Using the `mobsim` package from the DataWrangling module last week, simulate two local communities each containing 1000 individuals (*N*) and 25 species (*S*), but with one having a random spatial distribution and the other having a patchy spatial distribution. 
Take ten (10) subsamples from each site using the quadrat function and answer the following questions:


1) Compare the average pairwise similarity among subsamples in site 1 (random spatial distribution) to the average pairswise similarity among subsamples in site 2 (patchy spatial distribution). 
Use a t-test to determine whether compositional similarity was affected by the spatial distribution. 
Finally, compare the compositional similarity of site 1 and site 2 to the source community?

```{r}
package.list <- c('mobsim', 'knitr', 'vegan', 'tidyr', 'dplyr', 'ggplot2', 'formatR')
for (package in package.list) {
  if (!require(package, character.only = TRUE, quietly = TRUE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
}
comeve <- sim_poisson_community(s_pool = 25, n_sim = 1000, sad_type = "lnorm", 
        sad_coef = list("meanlog" = 2, "sdlog" = 1))
com_eve1 <- sample_quadrats(comeve, n_quadrats = 10, quadrat_area = 0.01, 
             method = "random", avoid_overlap = T)  
compat <- sim_thomas_community(s_pool = 25, n_sim = 1000, sad_type = "lnorm", 
        sad_coef = list("meanlog" = 2, "sdlog" = 1))
com_pat1 <- sample_quadrats(compat, n_quadrats = 10, quadrat_area = 0.01, 
             method = "random", avoid_overlap = T)  

all_sites <- bind_rows(com_eve1$spec_dat, com_pat1$spec_dat)
beta.w(all_sites, 1, 12, pairwise = TRUE)
beta.w(com_eve1$spec_dat)
beta.w(com_pat1$spec_dat)

t.test(com_eve1$spec_dat, com_pat1$spec_dat)

```

>Beta diversity between sites 1 and 2 of their repective simulatins is 0.75 which means they are less diverse than the regional pool. With the t-test I got a p-value of 0.6985 which means that they are not different in a statistically significant way. Though sites 1 and 2 are more similar to one another than they are to either of their original community simulations as both the community simulations are overall more diverse than 1 and 2. 

2) Create a cluster diagram or ordination using your simulated data. 
Are there any visual trends that would suggest a difference in composition between site 1 and site 2? Describe.
```{r} 



com_evec <- vegdist(com_eve1$spec_dat, method = "bray", binary = TRUE)
ceclus <- hclust(com_evec, method = "ward.D2")

plot(ceclus, main = "Simulation Clustering", 
     ylab= "Squared Bray-Curtis Distance")


```
> Site 1 and 2 are not particularly close, but are not within entirely different groups composition wise. 
=======

2) Create a cluster diagram or ordination using your simulated data. 
Are there any visual trends that would suggest a difference in composition between site 1 and site 2? Describe.

