---
title: '8\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: 'Eli Graber; Z620: Quantitative Biodiversity, Indiana University'
date: "r format(Sys.time(), %d %B, %Y)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the  worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, it is *imperative* that you **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your '8.BetaDiversity' folder.
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**8.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**8.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, April 23^rd^, 2021 before 09:00 AM**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/8.BetaDiversity*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/QB2021_Graber/2.Worksheets/8.BetaDiversity")
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting
package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies')
  for (package in package.list) {
    if (!require(package, character.only = TRUE, quietly = TRUE)) {
      install.packages(package)
      library(package, character.only = TRUE)
    }
  }
data(doubs)


```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}

quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))
fish <- doubs$fish
fish <- fish[-8, ]
adonis(fish ~ quality, method = "bray", permutations = 999)
indval <- multipatt(fish, cluster = quality, func = "IndVal.g", control = how(nperm=999))
summary(indval)
fish.rel <- decostand(fish, method = "total")
phi <- multipatt(fish.rel, cluster = quality, func = "r.g", control = how(nperm=999))
summary(phi)

```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: Every species seems to be impacted by the quality of the lake and this is true across the different analysis. This does seem to agree with our visualizations in which we indicated that Alal and Satr were indicators. 


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}

fish.dist <- vegdist(doubs$fish[-8,], method = "bray")
env.dist <- vegdist(scale(doubs$env[-8,]), method = "euclid")
mantel(fish.dist, env.dist)

```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***:  There is a strong correlation between fish diversity and the quality of the stream environment. It supports it because a larger diversity of fish would be sustained by higher quality environments. 

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}

env.chem <- as.matrix(doubs$env[-8, 5:11])
fish.db <- vegdist(fish, method = "bray")
doubs.dbrda <-dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))
ordiplot(doubs.dbrda.mod0)
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)
doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[, c(4,6,7)], perm = 999)
dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1]/
                            sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] /
                             sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) *100
par(mar = c(5, 5, 4, 4) +0.1)
plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1), ylim = c(-1.1, 2.7), 
     xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%", sep = ""), 
     ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)
axis(side = 1, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
axis(side = 2, labels = T, lwd.ticks =2, cex.axis = 1.2, las =1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
points(scores(doubs.dbrda, display = "wa"), pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"), labels = row.names(scores(doubs.dbrda, display = "wa")))

vectors <- scores(doubs.dbrda, display = "bp")

arrows(0, 0, vectors[,1], vectors[,2], lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[,2], pos = 3, labels = row.names(vectors))
axis(side = 3, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2, at = pretty(range(vectors[,1])) * 2, labels = pretty(range(vectors[,1])))
axis(side = 4,lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2, at = pretty(range(vectors[,2])) * 2, labels = pretty(range(vectors[,2])))
```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***: Oxygen levels, nitrogen levels, and bdo significantly contribute to the variation in the fish community structure. 

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova
env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]
rs <- rowSums(fish)/sum(fish)

doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)

doubs.pcnmw$values > 0

doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

plot(step.pcnm)

step.pcnm$anova

space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 + PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]
doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))
permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutation = 999)
permutest(doubs.total.space, permutations = 999)
doubs.varpart <- varpart(fish.db, env.mod, space.mod)
doubs.varpart
par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -3)



```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: 74% of the variation is explained by environment (13%), space (24%), and environment and space (37%) with environment and space combined having the greatest impact on fish distrubtion. 27% of the variation is not explained by either of these factors. These factors are clearly large contributors to the variations fish distribution. 

## SYNTHESIS

As in the previous worksheet, use the `mobsim` package from the DataWrangling module to simulate two local communities each containing 1000 individuals (*N*) and 25 species (*S*), but with one having a random spatial distribution and the other having a patchy spatial distribution. 
Take ten (10) subsamples from each site using the quadrat function and answer the following questions:
```{r}
package.list <- c('mobsim', 'knitr', 'vegan', 'tidyr', 'dplyr', 'ggplot2', 'formatR')
for (package in package.list) {
  if (!require(package, character.only = TRUE, quietly = TRUE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
}
comeve <- sim_poisson_community(s_pool = 25, n_sim = 1000, sad_type = "lnorm", 
        sad_coef = list("meanlog" = 2, "sdlog" = 1))
com_eve1 <- sample_quadrats(comeve, n_quadrats = 10, quadrat_area = 0.01, 
             method = "random", avoid_overlap = T)  
compat <- sim_thomas_community(s_pool = 25, n_sim = 1000, sad_type = "lnorm", 
        sad_coef = list("meanlog" = 2, "sdlog" = 1))
com_pat1 <- sample_quadrats(compat, n_quadrats = 10, quadrat_area = 0.01, 
             method = "random", avoid_overlap = T)  
```
1) Perform a PERMANOVA to test whether or not the spatial distribution of species affects species composition.
```{r}

all_sites <- bind_rows(com_eve1$spec_dat, com_pat1$spec_dat)
distrib <- c(rep("Tom", 10), rep("Pos", 10))
adonis(all_sites ~ distrib, method = "bray", permutations = 999)


```


> For my randomly generated communities it appears there is a notable affect of spatial distribution on species composition producing a p-value of 0.001 which is well within the standard range for statistical significance. 

2) Load the dataset you are using for your Team Project. 
Perform an ordination to visualize your dataset.
Using this ordination, develop some hypotheses relevant to $\beta$-diversity.
Use a statistic to test one of these hypotheses.
Succinctly explain the finding and its relevance to your system.
```{r}
setwd("~/GitHub/QB2021_Team3")
abund <- exp(`135FoodWebs`$Log.Abundance.)
abund.db <- vegdist(abund, method = "bray")

pcoa <- cmdscale(abund.db, eig = TRUE, k = 3)
explainvar1 <- round(pcoa$eig[1] / sum(pcoa$eig), 3) * 100
explainvar2 <- round(pcoa$eig[2] / sum(pcoa$eig), 3) * 100
explainvar3 <- round(pcoa$eig[3] / sum(pcoa$eig), 3) * 100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

par(mar = c(5, 5, 1, 2) + 0.1)

plot(pcoa$points[ ,1], pcoa$points[ ,2], ylim = c(-0.2, 0.7),
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

points(pcoa$points[,1], pcoa$points[,2],
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(pcoa$points[,1], pcoa$points[, 2],
     labels = row.names(pcoa$points)) 
```

> We hypothesis that from the initial ordination plot the differences in abundance will be clusted by the human development of the environment from which the samples were taken. 

```{r}

  model = lm(I(Log.Abundance.)~Ecosystem.Type.ID, data=`135FoodWebs`)
summary(model)
    
```
> Using an ANOVA we can see that there is a statisticaly significant impact of the ecosystem time on species abundance.  Ecosystem ID 2 is organic and 3 is intensive farming (though 4 is super intesive so I am not sure how we can actually take away from that), showing that some abundances seems to be less impacted by the ecosystem type. 